The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-11-27 script/en/chapters/paper_and_pencil.tex
Script[de+en]: Many small changes, e.g. Sage samples collected in files, use of iload; described how to deal with Sage notebook using interact; substituted href by url.

In the following section some classic ciphers are implemented using the open source computer
algebra system (CAS) \verb!Sage!. The code was tested with Sage version 4.2.
All ciphers are explained in chapter~\ref{Kapitel_PaperandPencil}
(``\nameref{Kapitel_PaperandPencil}'').

\noindent A first introduction to the CAS Sage can be found in the appendix \ref{s:appendix-using-sage}.

\noindent To make the sample code\footnote{%
     Further examples with Sage can be found e.g.\ at:\\
     \href{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}{\tt
           http://www.sagemath.org/doc/constructions/linear\_codes.html\#classical-ciphers}
                                          }
easier to understand, we used the structure and the naming conventions
shown in the graphics below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \begin{itemize}
    \item Encoding adapts the letters in the given plaintext P to the case defined in the
          given alphabet, and all non-alphabet characters are filtered out.
    \item Enciphering creates the ciphertext C.
  \end{itemize}
  \item Decryption also consists of two steps: deciphering and decoding.
  \begin{itemize}
    \item Decoding is only necessary if the symbols in the alphabet are not ASCII characters.
  \end{itemize}
\end{itemize}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14, code more variable
sage: keylen = 14
sage: shift = 2
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen)
sage:
sage: # construct the plaintext string from the first 14 letters of the alphabet plus blanks
sage: # plaintext   = "A B C D E F G H I J K L M N"
sage: A.gens()
(A, B, C, D, E, F, G, H, I, J, K, L, M, N, O, P, Q, R, S, T, U, V, W, X, Y, Z)
sage: P=''
sage: for i in range(keylen): P=P + " " + str(A.gen(i))
....:
sage: P
' A B C D E F G H I J K L M N'
sage:
sage: # encryption key
sage: # key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: key = [(i+shift).mod(keylen) + 1 for i in range(keylen)]; key
[3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in one step)
sage: C   = T.enciphering(key, msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: # keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) + 1 for i in range(keylen)]; keyInv
[13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: DC   = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMN
sage:
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
sage: # using the deciphering method requires to change the type of the variable key
sage: DC  = T.deciphering(T(key).key(), C); DC
ABCDEFGHIJKLMN
sage:
sage: # representation of key and inverse key as permutations
sage: T(key)
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(key).key()
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(keyInv)
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple Transposition by shifting (key and inverse key constructed with "range")}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: #### We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic Substitution with randomly generated key}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #### Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitly given, step-by-step approach)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #### Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; substitution keys are generated)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: shiftcipher = ShiftCryptosystem(AlphabeticStrings()); shiftcipher
Shift cryptosystem on Free alphabetic string monoid on A-Z
sage: P = shiftcipher.encoding("Shift me any number of positions."); P
SHIFTMEANYNUMBEROFPOSITIONS
sage: key = 12  # shift can be any integer number
sage: 
sage: # shift the plaintext by 12 positions to get the ciphertext
sage: C = shiftcipher.enciphering(key, P); C
ETURFYQMZKZGYNQDARBAEUFUAZE
sage: 
sage: # decrypt the ciphertext and ensure that it is the original plaintext
sage: DC = shiftcipher.deciphering(key, C); DC
SHIFTMEANYNUMBEROFPOSITIONS
sage: DC == P
True
\end{Verbatim}
\caption{A shift cipher over the upper-case letters of the English alphabet}
\label{paper_pencil:shift_cipher:Sage_example}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet
sage: A = BinaryStrings()
sage: # substitution cipher over the alphabet A; no keylen argument possible
sage: S = SubstitutionCryptosystem(A); S
Substitution cryptosystem on Free binary string monoid
sage: # to have a substitute for each symbol, key has always the length of the alphabet
sage: key = S.random_key(); key
10
sage: len(key)
2
sage: P = "Working with binary numbers."
sage: # encryption
sage: msg = A.encoding(P); msg
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001100010011010010110111001100001011100100111100100\
1000000110111001110101011011010110001001100101011100100111001100101110
sage: C   = S.enciphering(key, msg); C
10101000100100001000110110010100100101101001000110011000110111111000100010010\
11010001011100101111101111110011101100101101001000110011110100011011000011011\
0111111001000110001010100100101001110110011010100011011000110011010001
sage: # decryption
sage: DC  = S.deciphering(key, C); DC
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001100010011010010110111001100001011100100111100100\
1000000110111001110101011011010110001001100101011100100111001100101110
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with a binary alphabet}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = HexadecimalStrings()
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
2b56a4e701c98df3
sage: len(key)
16
sage: # Number of possible keys
sage: factorial(len(key))
20922789888000
sage: P   = "Working with a larger alphabet."
sage:
sage: msg = A.encoding(P); msg
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: C   = S.enciphering(key, msg); C
47e375e9e1efe75277e17ae052eb52e8eb75e7e47552ebe872e0ebe5e47a5f
sage: DC  = S.deciphering(key, C); DC
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: msg == DC
True
sage:
sage: # Conversion hex back to ASCII:
sage: # - AlphabeticStrings() and HexadecimalStrings() don't have according methods.
sage: # - So we used Python:
sage: #   - repr(DC) converts the StringMonoidElement DC to a string
sage: #     still containing hex symbols;
sage: #   - binascii.a2b_hex converts this hex string to an ASCII string.
sage: import binascii
sage: DDC = binascii.a2b_hex(repr(DC)); DDC
'Working with a larger alphabet.'
sage:
sage: P == DDC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with a hexadecimal alphabet (and decoding in Py\-thon)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Vigenere cipher
sage: keylen = 14
sage: A = AlphabeticStrings()
sage: V = VigenereCryptosystem(A, keylen); V
Vigenere cryptosystem on Free alphabetic string monoid on A-Z of period 14
sage:
sage: # alternative: given key: key = A('ABCDEFGHIJKLMN'); key
sage: key = V.random_key(); key
WSSSEEGVVAARUD
sage: len(key)
14
sage:
sage: # encoding
sage: P = "The Vigenere cipher is polyalphabetic."
sage: len(P)
38
sage: msg = V.encoding(P); msg     # alternative: msg = A.encoding(P); msg
THEVIGENERECIPHERISPOLYALPHABETIC
sage:
sage: # encryption [2 alternative ways (in two steps or in one): both work]
sage: # encrypt = V(key); encrypt
sage: # C = encrypt(msg); C
sage: C   = V.enciphering(key, msg); C
PZWNMKKIZRETCSDWJAWTUGTALGBDXWLAG
sage:
sage: # decryption
sage: DC  = V.deciphering(key, C); DC
THEVIGENERECIPHERISPOLYALPHABETIC
sage: msg == DC
True
\end{Verbatim}
\caption{Vigen{\`e}re cipher}
\end{sagecode}

@@

>>>>

In the following section some classic ciphers are implemented using the open source computer
algebra system \verb!Sage!\footnote{%
    A first introduction to the CAS Sage can be found in the appendix \ref{s:appendix-using-sage}.
                        }
. The code was tested with Sage version 4.2.
All ciphers are explained in chapter~\ref{Kapitel_PaperandPencil}
(``\nameref{Kapitel_PaperandPencil}'').

\noindent To make the sample code\footnote{%
     Further examples with Sage concerning classic crypto methods can be found e.g.:\\
     - at \url{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}\\
     - in the thesis of Minh Van Nguyen \cite{pp:Nguyen2009}
                                          }
easier to understand, we used the structure and the naming conventions
shown in the graphics below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \begin{itemize}
    \item Encoding adapts the letters in the given plaintext P to the case defined in the
          given alphabet, and all non-alphabet characters are filtered out.
    \item Enciphering creates the ciphertext C.
  \end{itemize}
  \item Decryption also consists of two steps: deciphering and decoding.
  \begin{itemize}
    \item Decoding is only necessary if the symbols in the alphabet are not ASCII characters.
  \end{itemize}
\end{itemize}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14, code more variable
sage: keylen = 14
sage: shift = 2
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen)
sage:
sage: # construct the plaintext string from the first 14 letters of the alphabet plus blanks
sage: # plaintext   = "A B C D E F G H I J K L M N"
sage: A.gens()
(A, B, C, D, E, F, G, H, I, J, K, L, M, N, O, P, Q, R, S, T, U, V, W, X, Y, Z)
sage: P=''
sage: for i in range(keylen): P=P + " " + str(A.gen(i))
....:
sage: P
' A B C D E F G H I J K L M N'
sage:
sage: # encryption key
sage: # key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: key = [(i+shift).mod(keylen) + 1 for i in range(keylen)]; key
[3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in one step)
sage: C   = T.enciphering(key, msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: # keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) + 1 for i in range(keylen)]; keyInv
[13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: DC   = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMN
sage:
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
sage: # using the deciphering method requires to change the type of the variable key
sage: DC  = T.deciphering(T(key).key(), C); DC
ABCDEFGHIJKLMN
sage:
sage: # representation of key and inverse key as permutations
sage: T(key)
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(key).key()
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(keyInv)
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple Transposition by shifting (key and inverse key constructed with (``range''}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: # We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic Substitution with randomly generated key}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: # Next, we recover the plaintext.
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitly given, step-by-step approach)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: # Next, we recover the plaintext.
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; substitution key generated)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Shift cipher directly
sage: shiftcipher = ShiftCryptosystem(AlphabeticStrings()); shiftcipher
Shift cryptosystem on Free alphabetic string monoid on A-Z
sage: P = shiftcipher.encoding("Shift me any number of positions."); P
SHIFTMEANYNUMBEROFPOSITIONS
sage: key = 12  # shift can be any integer number
sage: 
sage: # shift the plaintext by 12 positions to get the ciphertext
sage: C = shiftcipher.enciphering(key, P); C
ETURFYQMZKZGYNQDARBAEUFUAZE
sage: 
sage: # decrypt the ciphertext and ensure that it is the original plaintext
sage: DC = shiftcipher.deciphering(key, C); DC
SHIFTMEANYNUMBEROFPOSITIONS
sage: DC == P
True
\end{Verbatim}
\caption{A shift cipher over the upper-case letters of the English alphabet}
\label{paper_pencil:shift_cipher:Sage_example}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet
sage: B = BinaryStrings()
sage: # substitution cipher over the alphabet B; no keylen argument possible
sage: S = SubstitutionCryptosystem(B); S
Substitution cryptosystem on Free binary string monoid
sage: # To get a substitute for each symbol, key has always the length of the alphabet
sage: key = S.random_key(); key
10
sage: len(key)
2
sage: P = "Working with binary numbers."
sage: # encryption
sage: msg = B.encoding(P); msg
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001100010011010010110111001100001011100100111100100\
1000000110111001110101011011010110001001100101011100100111001100101110
sage: C   = S.enciphering(key, msg); C
10101000100100001000110110010100100101101001000110011000110111111000100010010\
11010001011100101111101111110011101100101101001000110011110100011011000011011\
0111111001000110001010100100101001110110011010100011011000110011010001
sage: # decryption
sage: DC  = S.deciphering(key, C); DC
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001100010011010010110111001100001011100100111100100\
1000000110111001110101011011010110001001100101011100100111001100101110
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with a binary alphabet}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = HexadecimalStrings()
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
2b56a4e701c98df3
sage: len(key)
16
sage: # Number of possible keys
sage: factorial(len(key))
20922789888000
sage: P   = "Working with a larger alphabet."
sage:
sage: msg = A.encoding(P); msg
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: C   = S.enciphering(key, msg); C
47e375e9e1efe75277e17ae052eb52e8eb75e7e47552ebe872e0ebe5e47a5f
sage: DC  = S.deciphering(key, C); DC
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: msg == DC
True
sage:
sage: # Conversion hex back to ASCII:
sage: # - AlphabeticStrings() and HexadecimalStrings() don't have according methods.
sage: # - So we used Python directly.
sage: import binascii
sage: DDC = binascii.a2b_hex(repr(DC)); DDC
'Working with a larger alphabet.'
sage:
sage: P == DDC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with a hexadecimal alphabet (and decoding in Py\-thon)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Vigenere cipher
sage: keylen = 14
sage: A = AlphabeticStrings()
sage: V = VigenereCryptosystem(A, keylen); V
Vigenere cryptosystem on Free alphabetic string monoid on A-Z of period 14
sage:
sage: # alternative could be a given key: key = A('ABCDEFGHIJKLMN'); key
sage: key = V.random_key(); key
WSSSEEGVVAARUD
sage: len(key)
14
sage:
sage: # encoding
sage: P = "The Vigenere cipher is polyalphabetic."
sage: len(P)
38
sage: msg = V.encoding(P); msg     # alternative: msg = A.encoding(P); msg
THEVIGENERECIPHERISPOLYALPHABETIC
sage:
sage: # encryption [2 alternative ways (in two steps or in one): both work]
sage: # encrypt = V(key); encrypt
sage: # C = encrypt(msg); C
sage: C   = V.enciphering(key, msg); C
PZWNMKKIZRETCSDWJAWTUGTALGBDXWLAG
sage:
sage: # decryption
sage: DC  = V.deciphering(key, C); DC
THEVIGENERECIPHERISPOLYALPHABETIC
sage: msg == DC
True
\end{Verbatim}
\caption{Vigen{\`e}re cipher}
\end{sagecode}

@@

\bibitem[Nguyen2009]{pp:Nguyen2009} \index{Nguyen 2009} 
       Minh Van Nguyen, \\
       {\em Exploring Cryptography Using the Sage Computer Algebra System}, \\
       Victoria University, 2009, \\
       See Sage publications page
       \url{http://www.sagemath.org/library-publications.html}
